---
title: "pannet: validation against classical estimators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pannet: validation against classical estimators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = requireNamespace("panglm", quietly = TRUE) && requireNamespace("plm", quietly = TRUE)
)
```

## Why this vignette exists

`pannet_benchmark()` reports predictive performance, but predictive RMSE can
look fine even when a model's *coefficients* are biased -- especially for a
model whose whole point is capturing nonlinear effects, where "the right
answer" isn't obvious from held-out error alone. This vignette checks
`pannet` against classical estimators (`panglm`, `plm`) on realistic panel
sizes: small N (~75), short T (6-8), sparse/zero-inflated counts, and a
genuine dynamic-panel DGP -- and reports what it finds, including where
`pannet` currently falls short, not just where it works.

```{r setup}
library(pannet)
library(panglm)
library(plm)
```

## 1. Small N, short T, linear DGP (the easy case)

```{r linear-recovery}
sim <- pannet_simulate(N = 75, T = 6, n_x = 2, dgp = "linear", family = "gaussian", seed = 1)
d <- sim$data
sim$true_beta

f_pannet <- pannet(y ~ x1 + x2, data = d, id = id, time = time, family = "gaussian",
                    model = "individual", epochs = 150, verbose = FALSE, seed = 1)
f_panglm <- panglm(y ~ x1 + x2, data = d, index = c("id", "time"), model = "within", family = "gaussian")

rbind(
  pannet  = c(marginal_effects(f_pannet, "x1")$marginal_effect,
              marginal_effects(f_pannet, "x2")$marginal_effect),
  panglm  = coef(f_panglm)
)
```

`panglm`'s exact within estimator recovers the true slopes closely (as it
must -- it's a closed-form solution to the correctly-specified linear model).
`pannet`'s marginal effects are in the right ballpark but noisier, which
section 4 below shows is largely a seed-to-seed instability issue rather
than a systematic bias for this DGP.

## 2. Sparse, zero-inflated count panel

A DGP where most units are structurally near-zero and a few are highly
active -- the kind of panel that first surfaced a singular-Hessian crash in
`panglm`'s random-effects Poisson fit (now fixed; see `panglm`'s `NEWS.md`).

```{r sparse-count}
set.seed(11)
N <- 75; Tt <- 6
dc <- data.frame(id = rep(1:N, each = Tt), time = rep(1:Tt, N))
dc$x1 <- rnorm(nrow(dc))
u <- rep(c(rep(-4, round(N * 0.85)), rep(1.5, N - round(N * 0.85))), each = Tt)
true_beta <- 0.4
dc$y <- rpois(nrow(dc), lambda = exp(-0.5 + true_beta * dc$x1 + u))
cat("Fraction zero:", mean(dc$y == 0), "\n")

f_pc <- pannet(y ~ x1, data = dc, id = id, time = time, family = "poisson",
                model = "individual", epochs = 200, verbose = FALSE, seed = 1)
f_gc <- panglm(y ~ x1, data = dc, index = c("id", "time"), model = "within", family = "poisson")

cat("true beta:            ", true_beta, "\n")
cat("panglm FE coef(x1):   ", coef(f_gc)["x1"], "\n")
cat("pannet AME(x1):       ", marginal_effects(f_pc, "x1")$marginal_effect, "\n")
```

**Finding:** on this DGP, `panglm`'s exact conditional-MLE estimator recovers
the true slope closely; `pannet`'s marginal effect is meaningfully biased
downward (by roughly a third). This bias is **not** a network-capacity/
overfitting artifact -- it persists essentially unchanged across
`hidden = c(64,32)` (the default), `hidden = c(16,8)`, `hidden = c(16,8)` with
dropout, and even `hidden = integer(0)` (a linear net, mathematically
equivalent to a GLM backbone). The likely cause is incidental-parameters-type
bias from jointly estimating one embedding per individual (75 of them) under
heavy sparsity, rather than profiling the individual effect out in closed
form the way `panglm`'s conditional MLE does. **Regularization is not the
fix here** -- this is a structural property of the dummy-variable/embedding
approach to fixed effects under sparsity, worth keeping in mind rather than
tuning around.

## 3. Dynamic panel

```{r dynamic-panel}
set.seed(5)
N <- 75; Tt <- 8
dd <- data.frame(id = rep(1:N, each = Tt), time = rep(1:Tt, N))
a_i <- rep(rnorm(N, sd = 1), each = Tt)
rho <- 0.6; beta_x <- 0.5
dd$x1 <- rnorm(nrow(dd))
y <- numeric(nrow(dd))
for (i in seq_len(N)) {
  idx <- which(dd$id == i)
  y_prev <- 0
  for (t in idx) {
    y[t] <- rho * y_prev + beta_x * dd$x1[t] + a_i[t] + rnorm(1, sd = 0.5)
    y_prev <- y[t]
  }
}
dd$y <- y

f_dyn <- pannet(y ~ x1, data = dd, id = id, time = time, family = "gaussian",
                 model = "dynamic", lags = 1, epochs = 300, verbose = FALSE, seed = 1)
cat("true beta_x:         ", beta_x, "  pannet AME(x1):    ", marginal_effects(f_dyn, "x1")$marginal_effect, "\n")
cat("true rho (lag coef): ", rho,    "  pannet AME(lag1_y):", marginal_effects(f_dyn, "lag1_y")$marginal_effect, "\n")
```

**Finding:** the covariate slope (`x1`) recovers reasonably. The lagged-
outcome coefficient is substantially overestimated. This matches the
well-known **Nickell (1981) dynamic panel bias**: *any* fixed-effects
estimator (not specific to `pannet`) is inconsistent for the lagged-dependent-
variable coefficient when both individual effects and a lagged `y` are
present with finite T -- correcting it requires a specialized estimator
(Arellano-Bond-style GMM), which neither `pannet` nor `panglm` currently
implements. Treat `pannet`'s dynamic-panel lag coefficients as unreliable
until such a correction exists; the covariate slopes are comparatively more
trustworthy.

## 4. Seed stability

```{r seed-stability}
sim2 <- pannet_simulate(N = 75, T = 6, n_x = 2, dgp = "linear", family = "gaussian", seed = 1)
d2 <- sim2$data
true_x1 <- sim2$true_beta["x1"]

ames <- sapply(1:5, function(s) {
  f <- pannet(y ~ x1 + x2, data = d2, id = id, time = time, family = "gaussian",
              model = "individual", epochs = 150, verbose = FALSE, seed = s)
  marginal_effects(f, "x1")$marginal_effect
})
cat("true beta(x1):", true_x1, "\n")
cat("AME(x1) across 5 seeds:", round(ames, 4), "\n")
cat("range:", diff(range(ames)), "\n")
```

**Finding:** for a small true effect, seed-to-seed variability in the
estimated marginal effect can be comparable to or larger than the true
effect itself. **Don't trust a single-seed `pannet` fit for a marginal
effect close to zero** -- refit across a few seeds (or use `pannet_simulate()`
to check recovery on comparable synthetic data first) and treat the spread
as part of the uncertainty, not just the reported point estimate.

## Summary

| Scenario | Recovers truth? |
|---|---|
| Small N/T, linear, gaussian | Roughly, with seed-dependent noise |
| Sparse/zero-inflated count | No -- biased downward, not fixed by regularization |
| Dynamic panel, covariate slope | Roughly |
| Dynamic panel, lag coefficient | No -- Nickell bias, not `pannet`-specific |

`pannet`'s comparative advantage is genuinely nonlinear covariate-response
shapes (see the main introduction vignette and the package's own
Monte Carlo simulation in the manuscript), not these edge cases. Where a
classical closed-form estimator like `panglm` applies and the true effect is
plausibly linear, prefer it -- it's both faster and, on this evidence, more
reliable in exactly the scenarios (sparse counts, dynamic panels, small true
effects) where `pannet` currently struggles.
