---
title: "pannet: validation against classical estimators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pannet: validation against classical estimators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = requireNamespace("panglm", quietly = TRUE) && requireNamespace("plm", quietly = TRUE)
)
```

## Why this vignette exists

`pannet_benchmark()` reports predictive performance, but predictive RMSE can
look fine even when a model's *coefficients* are biased -- especially for a
model whose whole point is capturing nonlinear effects, where "the right
answer" isn't obvious from held-out error alone. This vignette checks
`pannet` against classical estimators (`panglm`, `plm`) on realistic panel
sizes: small N (~75), short T (6-8), sparse/zero-inflated counts, and a
genuine dynamic-panel DGP -- and reports what it finds, including where
`pannet` currently falls short, not just where it works.

```{r setup}
library(pannet)
library(panglm)
library(plm)
```

## 1. Small N, short T, linear DGP (the easy case)

```{r linear-recovery}
sim <- pannet_simulate(N = 75, T = 6, n_x = 2, dgp = "linear", family = "gaussian", seed = 1)
d <- sim$data
sim$true_beta

f_pannet <- pannet(y ~ x1 + x2, data = d, id = id, time = time, family = "gaussian",
                    model = "individual", epochs = 150, verbose = FALSE, seed = 1)
f_panglm <- panglm(y ~ x1 + x2, data = d, index = c("id", "time"), model = "within", family = "gaussian")

rbind(
  pannet  = c(marginal_effects(f_pannet, "x1")$marginal_effect,
              marginal_effects(f_pannet, "x2")$marginal_effect),
  panglm  = coef(f_panglm)
)
```

`panglm`'s exact within estimator recovers the true slopes closely (as it
must -- it's a closed-form solution to the correctly-specified linear model).
`pannet`'s marginal effects are in the right ballpark but noisier, which
section 4 below shows is largely a seed-to-seed instability issue rather
than a systematic bias for this DGP.

## 2. Sparse, zero-inflated count panel

A DGP where most units are structurally near-zero and a few are highly
active -- the kind of panel that first surfaced a singular-Hessian crash in
`panglm`'s random-effects Poisson fit (now fixed; see `panglm`'s `NEWS.md`).

```{r sparse-count}
set.seed(11)
N <- 75; Tt <- 6
dc <- data.frame(id = rep(1:N, each = Tt), time = rep(1:Tt, N))
dc$x1 <- rnorm(nrow(dc))
u <- rep(c(rep(-4, round(N * 0.85)), rep(1.5, N - round(N * 0.85))), each = Tt)
true_beta <- 0.4
dc$y <- rpois(nrow(dc), lambda = exp(-0.5 + true_beta * dc$x1 + u))
cat("Fraction zero:", mean(dc$y == 0), "\n")

f_pc <- pannet(y ~ x1, data = dc, id = id, time = time, family = "poisson",
                model = "individual", epochs = 200, verbose = FALSE, seed = 1)
f_gc <- panglm(y ~ x1, data = dc, index = c("id", "time"), model = "within", family = "poisson")

cat("true beta:            ", true_beta, "\n")
cat("panglm FE coef(x1):   ", coef(f_gc)["x1"], "\n")
cat("pannet AME(x1):       ", marginal_effects(f_pc, "x1")$marginal_effect, "\n")
```

**Finding:** on this DGP, `panglm`'s exact conditional-MLE estimator recovers
the true slope closely; `pannet`'s marginal effect is meaningfully biased
downward (by roughly a third). This bias is **not** a network-capacity/
overfitting artifact -- it persists essentially unchanged across
`hidden = c(64,32)` (the default), `hidden = c(16,8)`, `hidden = c(16,8)` with
dropout, and even `hidden = integer(0)` (a linear net, mathematically
equivalent to a GLM backbone). The likely cause is incidental-parameters-type
bias from jointly estimating one embedding per individual (75 of them) under
heavy sparsity, rather than profiling the individual effect out in closed
form the way `panglm`'s conditional MLE does. **Regularization is not the
fix here** -- this is a structural property of the dummy-variable/embedding
approach to fixed effects under sparsity, worth keeping in mind rather than
tuning around.

## 3. Dynamic panel

```{r dynamic-panel}
set.seed(5)
N <- 75; Tt <- 8
dd <- data.frame(id = rep(1:N, each = Tt), time = rep(1:Tt, N))
a_i <- rep(rnorm(N, sd = 1), each = Tt)
rho <- 0.6; beta_x <- 0.5
dd$x1 <- rnorm(nrow(dd))
y <- numeric(nrow(dd))
for (i in seq_len(N)) {
  idx <- which(dd$id == i)
  y_prev <- 0
  for (t in idx) {
    y[t] <- rho * y_prev + beta_x * dd$x1[t] + a_i[t] + rnorm(1, sd = 0.5)
    y_prev <- y[t]
  }
}
dd$y <- y

f_dyn <- pannet(y ~ x1, data = dd, id = id, time = time, family = "gaussian",
                 model = "dynamic", lags = 1, epochs = 300, verbose = FALSE, seed = 1)
cat("true beta_x:         ", beta_x, "  pannet AME(x1):    ", marginal_effects(f_dyn, "x1")$marginal_effect, "\n")
cat("true rho (lag coef): ", rho,    "  pannet AME(lag1_y):", marginal_effects(f_dyn, "lag1_y")$marginal_effect, "\n")
```

**Finding:** the covariate slope (`x1`) recovers reasonably. The lagged-
outcome coefficient is substantially overestimated. This matches the
well-known **Nickell (1981) dynamic panel bias**: *any* fixed-effects
estimator (not specific to `pannet`) is inconsistent for the lagged-dependent-
variable coefficient when both individual effects and a lagged `y` are
present with finite T -- correcting it requires a specialized estimator
(Arellano-Bond-style GMM), which neither `pannet` nor `panglm` currently
implements. Treat `pannet`'s dynamic-panel lag coefficients as unreliable
until such a correction exists; the covariate slopes are comparatively more
trustworthy.

## 4. Seed stability

```{r seed-stability}
sim2 <- pannet_simulate(N = 75, T = 6, n_x = 2, dgp = "linear", family = "gaussian", seed = 1)
d2 <- sim2$data
true_x1 <- sim2$true_beta["x1"]

ames <- sapply(1:5, function(s) {
  f <- pannet(y ~ x1 + x2, data = d2, id = id, time = time, family = "gaussian",
              model = "individual", epochs = 150, verbose = FALSE, seed = s)
  marginal_effects(f, "x1")$marginal_effect
})
cat("true beta(x1):", true_x1, "\n")
cat("AME(x1) across 5 seeds:", round(ames, 4), "\n")
cat("range:", diff(range(ames)), "\n")
```

**Finding:** for a small true effect, seed-to-seed variability in the
estimated marginal effect can be comparable to or larger than the true
effect itself. **Don't trust a single-seed `pannet` fit for a marginal
effect close to zero** -- refit across a few seeds (or use `pannet_simulate()`
to check recovery on comparable synthetic data first) and treat the spread
as part of the uncertainty, not just the reported point estimate.

## 5. Genuine multi-step forecasting: nonlinear dynamics

The dynamic-panel finding above only tested one-step-ahead coefficient
recovery. The more direct question for a *forecasting* use case is: does
`pannet_forecast()`'s recursive multi-step forecast beat a linear AR(1)+FE
benchmark when the true dynamics are genuinely nonlinear? A threshold
autoregressive (TAR) DGP -- fast mean-reversion when the outcome is
positive, near-unit-root persistence when negative -- is a standard
nonlinearity a single linear AR coefficient cannot represent.

```{r tar-forecast}
set.seed(42)
N <- 75; Tt_train <- 10; Tt_fore <- 4
simulate_unit <- function(n_periods, x, a_i, y0 = 0) {
  y <- numeric(n_periods); yp <- y0
  for (t in seq_len(n_periods)) {
    g <- if (yp < 0) 0.85 * yp else 0.2 * yp
    yp <- g + 0.5 * x[t] + a_i + rnorm(1, sd = 0.3)
    y[t] <- yp
  }
  y
}
a_i <- rnorm(N, sd = 0.5)
x_all <- matrix(rnorm(N * (Tt_train + Tt_fore)), nrow = N)
train_list <- vector("list", N); future_true <- vector("list", N)
for (i in seq_len(N)) {
  y_full <- simulate_unit(Tt_train + Tt_fore, x_all[i, ], a_i[i])
  train_list[[i]] <- data.frame(id = i, time = seq_len(Tt_train), x1 = x_all[i, 1:Tt_train], y = y_full[1:Tt_train])
  future_true[[i]] <- data.frame(id = i, time = Tt_train + seq_len(Tt_fore),
                                  x1 = x_all[i, (Tt_train+1):(Tt_train+Tt_fore)],
                                  y_true = y_full[(Tt_train+1):(Tt_train+Tt_fore)])
}
d_train <- do.call(rbind, train_list); d_future <- do.call(rbind, future_true)

f_nn <- pannet(y ~ x1, data = d_train, id = id, time = time, family = "gaussian",
               model = "dynamic", lags = 1, hidden = c(32L, 16L), epochs = 300,
               verbose = FALSE, seed = 1)
fc_nn <- pannet_forecast(f_nn, d_future[, c("id", "time", "x1")])
merged <- merge(d_future, fc_nn, by = c("id", "time"))
cat("pannet (nonlinear, recursive) overall RMSE:", sqrt(mean((merged$y_true - merged$forecast)^2)), "\n")
```

A linear AR(1)+FE benchmark, forecast the same way (recursively, feeding
its own forecasts back in as the next lag):

```{r tar-linear-benchmark}
d_train$id_f <- factor(d_train$id)
d_train_lag <- do.call(rbind, lapply(unique(d_train$id), function(i) {
  s <- d_train[d_train$id == i, ]; s <- s[order(s$time), ]
  s$lag1_y <- c(NA, s$y[-nrow(s)]); s
}))
lm_fit <- lm(y ~ x1 + lag1_y + id_f, data = d_train_lag)
last_obs <- do.call(rbind, lapply(split(d_train_lag, d_train_lag$id), function(s) s[which.max(s$time), ]))
hist_y <- setNames(last_obs$y, last_obs$id)
out <- list()
for (ft in sort(unique(d_future$time))) {
  rows <- d_future[d_future$time == ft, ]
  rows$lag1_y <- hist_y[as.character(rows$id)]
  rows$id_f <- factor(rows$id, levels = levels(d_train_lag$id_f))
  pred <- predict(lm_fit, newdata = rows)
  hist_y[as.character(rows$id)] <- pred
  out[[as.character(ft)]] <- data.frame(id = rows$id, time = ft, forecast = pred)
}
fc_lm <- do.call(rbind, out)
merged_lm <- merge(d_future, fc_lm, by = c("id", "time"))
cat("linear AR(1)+FE overall RMSE:", sqrt(mean((merged_lm$y_true - merged_lm$forecast)^2)), "\n")
```

**Finding:** on this DGP, at a realistic short-T panel (T=10 training
periods), the **linear AR(1)+FE benchmark forecasts better than `pannet`
at every horizon**, and the gap widens with the forecast horizon (errors
compound faster for `pannet`'s recursive forecast). This held up even
after fixing a real bug that had been silently zeroing out the lagged
history (see `NEWS.md`) -- it is not an artifact of that bug.

Repeating the same comparison with T=40 training periods per unit (4x more
data, same DGP) narrows the gap substantially (from ~40% worse to ~10%
worse), consistent with a bias-variance explanation: reliably estimating a
genuine nonlinear (threshold) shape needs more data than a single global
linear slope does, and multi-step recursive forecasting compounds whatever
extra estimation variance the nonlinear model carries. **Don't assume
`pannet`'s recursive forecast beats a linear dynamic benchmark just because
the true DGP is nonlinear -- check both, especially on short panels.** The
nonlinear-dynamics advantage this package is built around likely needs
more periods per unit than T~10 to show up in *forecasts* specifically
(as opposed to in-sample coefficient/shape recovery, which is a different
question -- see the main introduction vignette and the package's own
Monte Carlo simulation).

## Summary

| Scenario | Recovers truth? |
|---|---|
| Small N/T, linear, gaussian | Roughly, with seed-dependent noise |
| Sparse/zero-inflated count | No -- biased downward, not fixed by regularization |
| Dynamic panel, covariate slope | Roughly |
| Dynamic panel, lag coefficient | No -- Nickell bias, not `pannet`-specific |
| Nonlinear dynamics, multi-step forecast (T=10) | No -- linear AR+FE forecasts better |
| Nonlinear dynamics, multi-step forecast (T=40) | Closer, gap narrows to ~10% |

`pannet`'s comparative advantage is genuinely nonlinear covariate-response
shapes (see the main introduction vignette and the package's own
Monte Carlo simulation in the manuscript), not these edge cases. Where a
classical closed-form estimator like `panglm` applies and the true effect is
plausibly linear, prefer it -- it's both faster and, on this evidence, more
reliable in exactly the scenarios (sparse counts, dynamic panels, small true
effects) where `pannet` currently struggles.
