---
title: "Introduction to pannet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to pannet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>")
library(pannet)
set.seed(1)
```

`pannet` fits neural panel regression models -- a multilayer perceptron
backbone with additive learnable individual/time effects -- trained natively
via `torch` (no external learner dependency). The wrapper handles panel-aware
preprocessing, `offset()` terms, and dynamic lag features.

## Which extractors work with which `model`

| Function | pooled | individual | time | twoway | dynamic |
|---|---|---|---|---|---|
| `individual_effects()` | `NULL` | yes | `NULL` | yes | `NULL` |
| `time_effects()` | `NULL` | `NULL` | yes | yes | `NULL` |
| `marginal_effects()` | yes | yes | yes | yes | yes |
| `predict()` | yes | yes | yes | yes | yes (known units only for lag history) |

`individual_effects()`/`time_effects()` return `NULL` (not an error) when the
fitted model has no corresponding embedding -- e.g. `model = "dynamic"` has
neither, since it only uses lagged history, not panel-effect embeddings.
Check `fit$model` if unsure which applies.

## Gaussian pooled model

```{r}
df <- data.frame(
  id = rep(1:5, each = 4),
  time = rep(1:4, times = 5),
  x1 = rnorm(20),
  x2 = runif(20),
  y = rnorm(20)
)

fit <- pannet(
  y ~ x1 + x2,
  data = df,
  id = id,
  time = time,
  family = "gaussian",
  model = "pooled",
  epochs = 1,
  verbose = FALSE,
  validation = 0.2
)

predict(fit, df)[1:3]
```

## Dynamic prediction

Dynamic fits can use training history when predicting new periods for known
units.

```{r}
fit_dyn <- pannet(
  y ~ x1,
  data = df,
  id = id,
  time = time,
  family = "gaussian",
  model = "dynamic",
  lags = 1,
  epochs = 1,
  verbose = FALSE,
  validation = 0.2
)

new_period <- data.frame(
  id = c(1, 2),
  time = c(5, 5),
  x1 = rnorm(2)
)

predict(fit_dyn, new_period)
```

## Multiclass and fractional outputs

```{r}
df_multi <- data.frame(
  id = rep(1:4, each = 5),
  time = rep(1:5, times = 4),
  x1 = rnorm(20),
  x2 = runif(20),
  y = factor(sample(letters[1:3], 20, TRUE))
)

fit_multi <- pannet(
  y ~ x1 + x2,
  data = df_multi,
  id = id,
  time = time,
  family = "multiclass",
  model = "pooled",
  epochs = 1,
  verbose = FALSE,
  validation = 0.2
)

predict(fit_multi, df_multi, type = "prob")[1:3, ]
```

```{r}
df_frac <- data.frame(
  id = rep(1:4, each = 5),
  time = rep(1:5, times = 4),
  x1 = rnorm(20),
  y = runif(20)
)

fit_frac <- pannet(
  y ~ x1,
  data = df_frac,
  id = id,
  time = time,
  family = "fractional",
  model = "pooled",
  epochs = 1,
  verbose = FALSE,
  validation = 0.2
)

marginal_effects(fit_frac, "x1")
```
